---
title: "Session 6 - Time series"
author: "Jessica De Rongé & Paul Servais"
format: revealjs
editor: visual
execute:
  echo: TRUE
  eval: TRUE
---

## Today

-   Time series

-   Multi-level model

-   Exercises

-   Q&A research paper

------------------------------------------------------------------------

# Time series

------------------------------------------------------------------------

## What is a time series?

-   Repeated observations

-   Ordered by time

-   One observation per time unit

------------------------------------------------------------------------

## Stationary

Stationary means no trend in the time series

```{r}
library(rio)
births <- import("http://robjhyndman.com/tsdldata/data/nybirths.dat")
births_ts <- ts(births, frequency=12, start=c(1946,1))
plot.ts(births_ts)

```

------------------------------------------------------------------------

## Stationary II

```{r}
library(tseries)
adf.test(births_ts)
```

------------------------------------------------------------------------

## Dependence

Relationship between an observation at time t and an observation at time t-h

Different processes:

-   White Noise

-   Moving Average (MA)

-   Autoregression (AR)

-   Integration (example, Random Walk)

------------------------------------------------------------------------

## Identification

```{r}
par(mfrow=c(1,2))
dt<-arima.sim(model=list(order=c(1,0,0),ar=0.7),n=100)
acf(dt, lag.max=10)
pacf(dt, lag.max=10)
```

------------------------------------------------------------------------

## Estimation

`arima(time series, order(p,d,q))`:

-   time series: name of your time series object

-   order: specify the ARIMA elements

    -   p: order of the autoregression
    -   d: order of the integration
    -   q: order of the moving average

------------------------------------------------------------------------

## Estimation II

```{r}
model<-arima(dt,order=c(1,0,0))
model
```

------------------------------------------------------------------------

## Diagnostic

Check that the residuals of your model are white noise

```{r}
plot.ts(model$residuals)
Box.test(model$residuals,lag=20, type="Ljung-Box")
```

------------------------------------------------------------------------

# Bivariate time series

------------------------------------------------------------------------

## Bivariate time series

Analysis of two time series

Evaluates the causality between the two variables:

-   Prewhitening

-   Granger causality

-   Error correction models of cointegration (see Jan Rovny notes)

------------------------------------------------------------------------

## Data

```{r}
library(vdemdata)
data("vdem")

demo <- subset(vdem, country_name == "Türkiye" & year>1988 & year < 2020)

library(readr)
ucdp <- read_csv("https://ucdp.uu.se/downloads/ged/ged201.csv")

conf <- subset(ucdp, country == "Turkey")
conf_year <- aggregate(best ~ year, data = conf, FUN = sum)

df <- merge(demo, conf_year, by = "year", all.x = TRUE)
df$best[is.na(df$best)] <- 0   # no conflict = 0

#v2x_libdem = democracy
#best = conflict deaths

```

------------------------------------------------------------------------

## Plotting time series

```{r}
dem_index<-ts(df$v2x_libdem, start = min(df$year), frequency = 1)
plot.ts(dem_index)
```

------------------------------------------------------------------------

## Plotting time series

```{r}
conflict<-ts(df$best, start = min(df$year), frequency = 1)
plot.ts(conflict)
```

------------------------------------------------------------------------

## Prewhitening

Start by doing the identification, estimation and diagnostic check.

Once this is done, you can check the cross-correlation of the residuals of the two time series

------------------------------------------------------------------------

## Identification {.scrollable}

```{r}
library(tseries)
adf.test(dem_index)
par(mfrow=c(1,2))
acf(dem_index, lag.max = 10)
pacf(dem_index, lag.max= 10)
```

```{r}
adf.test(conflict)
par(mfrow=c(1,2))
acf(conflict, lag.max = 10)
pacf(conflict, lag.max = 10)
```

------------------------------------------------------------------------

## Estimation {.scrollable}

```{r}
model1<-arima(dem_index,order=c(1,1,0))
model1
model2<-arima(conflict, order=c(1,1,0))
model2
```

------------------------------------------------------------------------

## Diagnostic {.scrollable}

```{r}
Box.test(model1$residuals ,lag=20,type="Ljung-Box")
Box.test(model2$residuals ,lag=20,type="Ljung-Box")
```

------------------------------------------------------------------------

### Cross-correlation

```{r}
dem_res<-model1$residuals
conflict_res<-model2$residuals
ccf(dem_res,conflict_res, lag.max=10)
```

------------------------------------------------------------------------

## Granger causality

`grangertest(DV ~ IV, order=p, data)`:

-   DV: dependent variable

-   IV: independent variable

-   order = p: p is the number of lags you want to take into account in the regression

-   data: name of your dataset

------------------------------------------------------------------------

```{r}
library(lmtest)
grangertest(best~v2x_libdem,order=4, data=df)
grangertest(v2x_libdem~best, order=4,data=df)
```

------------------------------------------------------------------------

# TSCS

------------------------------------------------------------------------

## TSCS

Times series for more than one unit

Problems of bias estimation with OLS:

-   Unit effects

-   Standard errors estimation

------------------------------------------------------------------------

## Data

```{r}
library(WDI)
library(vdemdata)
library(readr)
library(dplyr)

countries <- c("Turkey", "Nigeria", "Colombia", "Philippines", "Thailand")  # example set

# --- Democracy (V-Dem) ---
data("vdem")
demo <- vdem %>%
  filter(country_name %in% countries) %>%
  select(country_name, year, v2x_libdem)


# --- Conflict (UCDP GED summary by country-year) ---
ucdp <- read_csv("https://ucdp.uu.se/downloads/ged/ged201.csv")

conf <- ucdp %>%
  filter(country %in% countries) %>%
  group_by(country, year) %>%
  summarise(conflict_deaths = sum(best, na.rm = TRUE), .groups = "drop")

df <- demo %>%
  left_join(conf, by = c("country_name" = "country", "year" = "year"))

df$conflict_deaths[is.na(df$conflict_deaths)] <- 0

df <- df %>%
  filter(year > 1960)

```

------------------------------------------------------------------------

## Unit effects {.scrollable}

You need to check for unit effects before conducting your analysis.

You can do an ANOVA test.

If you have unit effects, you can use fixed or random effects

```{r}
library(plm)
data<-pdata.frame(df,index=c("country_name","year"))
coplot(conflict_deaths~year|country_name, type="l",data=data)
boxplot(conflict_deaths~country_name, data=data)
anova<-aov(conflict_deaths~country_name, data=data)
summary(anova)
```

------------------------------------------------------------------------

## Fixed or Random effects

`plm(formula, data, index=c("unit","time"), model)`:

-   formula: Y\~X

-   data: name of the dataset

-   index= c("unit","time"): specify the name of the unit and time variable

-   model: "within" for fixed effects estimator and "random" for random effects

------------------------------------------------------------------------

### Example fixed effects I

```{r}
fe<-plm(conflict_deaths~v2x_libdem,
        data=data,
        index=c("country_name","year"),
        model="within")

summary(fe)


```

------------------------------------------------------------------------

### Example fixed effects II

```{r}
fe2<-lm(conflict_deaths~v2x_libdem + factor(country_name)-1,
        data=data)
summary(fe2)
```

------------------------------------------------------------------------

### Example random effects

```{r}
re<-plm(conflict_deaths~v2x_libdem,
        data=data,
        index=c("country_name","year"),
        model="random")

summary(re)
```

------------------------------------------------------------------------

### Comparing FE and RE

Should you choose fixed or random effects? You can do a Hausman test if you do not know which one to choose.

```{r}
phtest(fe,re)
```

------------------------------------------------------------------------

## Panel Corrected Standard Errors

With TSCS, you also have a lot of problems with SE.

One solution proposed is to use Panel Corrected Standard Errors on OLS.

`pcse(model, groupN, groupT)`:

-   model: name of the lm object

-   groupN: name of the unit variable

-   groupT: name of the time variable

------------------------------------------------------------------------

## Example

```{r}
library(pcse)
ols<-lm(conflict_deaths~v2x_libdem,
        data=data)
pcse<-pcse(ols,groupN=data$country_name,groupT=data$year)
summary(pcse)
```

------------------------------------------------------------------------

## PCSE and LDV

Using PCSE only deals with the SE errors but does not take into account unit effects.

One solution is to add a lag dependent variable into the OLS which will capture the history of the dependent variable into the model and thus take into account some of the unit effects.

In R, you need to compute first the lagged dependent variable before the OLS and PCSE. You can use the `lag()` function.

------------------------------------------------------------------------

## PCSE and LDV II

```{r}
data<-data %>%
  group_by(country_name) %>%
  mutate(l.conflict_death=dplyr::lag(conflict_deaths))

data2<-na.omit(data)

ols2<-lm(conflict_deaths~l.conflict_death+v2x_libdem,
        data=data2)
pcse2<-pcse(ols2,groupN=data2$country_name,groupT=data2$year)
summary(pcse2)

```

------------------------------------------------------------------------

# Multi-level model {.scrollable}

Data with nested observations in larger unit

Incorporate information at the larger unit into the regression

`(1|variable)` to incorporate the effect

```{r}
# I am simulating data, so a seed will guarantee the same simulated data every time
set.seed(1312) 

data <- tibble(
  region = rep(1:10, each = 100),
  # 10 regions, 100 individuals each
  individual_income = rnorm(1000, mean = 50000, sd = 10000),
  # some control variables
  education_level = rnorm(1000, mean = 16, sd = 2),
  gender = sample(c("Male", "Female"), 1000, replace = TRUE),
  # a random and fun control variable
  number_of_pets = rpois(1000, lambda = 2)
) |>
  mutate(
    individual_income = scale(individual_income),
    education_level = scale(education_level),
    # cheating myself to significant results for regional variable
    regional_effect = rnorm(10, mean = 0, sd = 2)[region]  # greater SD for more variability
  ) |>
  mutate(voted = as.integer(
    runif(1000) < plogis(
      0.5 + 0.25 * individual_income + 0.15 * education_level +
        0.2 * (gender == "Female") + 0.1 * number_of_pets +
        regional_effect
    )
  )
  )
```

```{r}
# Fit a logistic mixed model
library(lme4)
model <-
  glmer(
    voted ~ individual_income + education_level + gender + number_of_pets +
      (1 | region),
    family = binomial,
    data = data
  )

summary(model)
```

------------------------------------------------------------------------

# Q&A

------------------------------------------------------------------------

# Exercises
