---
title: "Session 3: Multinomial Regression"
author: "Jessica De Rongé & Paul Servais"
format: revealjs
execute:
  echo: true
editor: visual
---

## Last session

Logistic regression

Odds

Homework

------------------------------------------------------------------------

## Homework

Comments?

Feedback?

------------------------------------------------------------------------

## This week

1.  Multinomial Regression
2.  How to write a research paper?
3.  Exercises

------------------------------------------------------------------------

## Multinomial Regression

What is a multinomial regression?

When should you use a multinomial regression?

------------------------------------------------------------------------

## Multinomial Regression

Similar to logistic regression

The dependent variable = categorical variable (K\>2)

Run simultaneously different simple logistic regressions

Always the same baseline for the different regressions

------------------------------------------------------------------------

## Multinommial Regression formula

-   Dependent variable Y (Y=a,...,K)

-   Independent variable X

\begin{align}
ln \left( \frac{Pr(Y=A|x)}{Pr(Y=K|x)} \right) &=\beta_{0,A} + \beta_{1,A}X \\
ln \left( \frac{Pr(Y=B|x)}{Pr(Y=K|x)} \right) &=\beta_{0,B} + \beta_{1,B}X \\
&\vdots\\
ln \left( \frac{Pr(Y=K-1|x)}{Pr(Y=K|x)} \right) &=\beta_{0,K-1} + \beta_{1,K-1}X \\
\end{align}

------------------------------------------------------------------------

## In R

In R, you will need to use the `multinom` function from the `nnet` package. In terms of argument, you need to specify three elements.

`multinom(formula, data, Hess)`:

-   formula: specify your equation as Y \~ X

-   data: specify the name of the dataset

-   Hess=TRUE: specify you want the Hessian matrix

------------------------------------------------------------------------

## Example

::: r-fit-text
In this example, we will use the ESS dataset from round 11 ([ESS11](https://github.com/scpo-quantimethods/quantitative_methods_two/raw/main/sessions/session3/data/ESS11e04_1.csv)). The objective of this analysis is to understand which predictors can help us understand party identification. We are interested in the Spanish case. We propose one simple hypothesis:

> **H1**: The socio-economic status of a citizen influences his party identification

```{r}
library(rio)
library(tidyverse)
ess<-import("https://github.com/scpo-quantimethods/quantitative_methods_two/raw/main/sessions/session3/data/ESS11e04_1.csv")
#ess<-read.csv("ESS11e04_1.csv")
data <-ess |> #this is a pipe you can also use %>%
  select(cntry, #select your variables of interest
         prtclhes,
         gndr,
         agea,
         eduyrs) |>
  filter(cntry=="ES")
```
:::

------------------------------------------------------------------------

## Data cleaning

```{r}
data_clean <- data |>
  mutate(across(c("prtclhes","eduyrs"), ~case_when(
      . %in% c(66, 77, 88, 99, 7777, 8888, 9999) ~ NA_integer_,
      TRUE ~ .
    )),
    agea=case_when(agea==999~NA_integer_,
                   TRUE~agea),
  gndr = recode(gndr, `1` = "Male", `2` = "Female"))

data_clean$party[data_clean$prtclhes==1]<-"PP"
data_clean$party[data_clean$prtclhes==2]<-"PSOE"
data_clean$party[data_clean$prtclhes==3]<-"Vox"
data_clean$party[data_clean$prtclhes==4]<-"SUMAR"
data_clean$party[data_clean$prtclhes==5]<-"ERC"
data_clean$party[data_clean$prtclhes==6]<-"JuntsxCat"
data_clean$party[data_clean$prtclhes==7]<-"EH-Bildu"
data_clean$party[data_clean$prtclhes==8]<-"EAJ-PNV"
data_clean$party[data_clean$prtclhes==9]<-"BNG"
data_clean$party[data_clean$prtclhes==10]<-"Coalición Canaria"
data_clean$party[data_clean$prtclhes==11]<-"UPN"
data_clean$party[data_clean$prtclhes==12]<-"PACMA"
data_clean$party[data_clean$prtclhes==13]<-"Unidas Podemos"
data_clean$party[data_clean$prtclhes==14]<-"Izquierda Unida"
data_clean$party[data_clean$prtclhes==16]<-"Cuidadanos"
data_clean$party[data_clean$prtclhes==17]<-"CUP"
data_clean$party[data_clean$prtclhes==18]<-"Compromís"
data_clean$party[data_clean$prtclhes==19]<-"Drago Canarias"
data_clean$party[data_clean$prtclhes==20]<-"Alianza Catalana"
data_clean$party[data_clean$prtclhes==50]<-"Other"
```

------------------------------------------------------------------------

## Descriptive statistics

:::: r-fit-text
::: panel-tabset
### Proportional table

```{r}
table(data_clean$gndr)
table(data_clean$party)

```

### bar-plot

```{r}
ggplot(data_clean,aes(x=gndr))+geom_bar()+theme_bw()
```

```{r}
library(ggplot2)
ggplot(data_clean,aes(x=party))+geom_bar()+coord_flip()+theme_bw()
```

### Descriptive statistics table

```{r}
data_clean |> select(agea,
                     eduyrs)|> summary()
```
:::
::::

------------------------------------------------------------------------

## Model

```{r}
model_data<-subset(data_clean,party%in% c("Vox","SUMAR","PSOE","PP"))
model_data$party<-as.factor(model_data$party)

library(nnet)
model1<-multinom(party~agea+eduyrs+gndr,
                data=model_data,
                Hess=TRUE)
```

------------------------------------------------------------------------

## Log-odds coefficients

```{r}
summary(model1)
```

------------------------------------------------------------------------

## Log-odds coefficients II

```{r}
library(stargazer)
stargazer(model1,type="text",
          covariate.labels=c("age","education years","Male"))
```

------------------------------------------------------------------------

## Odds

```{r}
exp(coef(model1))
```

------------------------------------------------------------------------

## Odds II {.scrollable}

```{r}
library(modelsummary)
modelsummary(model1,
             stars=TRUE,
             stars_method="simple",
             coef_map=c("agea"="Age",
                        "eduyrs"="Years of education",
                        "gndrMale"="Male"),
             title="Multinomial Regression Results",
             output="html",
             shape= model+term~response,
             gof_map=tibble::tribble(~raw,~clean,~fmt,
                                     "AIC","Akaike Inf. Crit.",2,
                                     "BIC","Bayesian Inf. Crit.",2,
                                     "logLik","Log-Likelihood",2,
                                     "deviance","Deviance",2),
             exponentiate = TRUE
             ) 
```

------------------------------------------------------------------------

## Interpretation

Which party is used as the baseline/reference?

::: notes
Which party is used as the baseline/reference? PP party

Which variables have statistically significant coefficients? age for SUMAR and VOX education years for SUMAR gender for PSOE and VOX

Do all variables have significant coefficients for all parties? no PSOE only gender SUMAR age and education Vox age and gender

What is the effect of the variable gender (in odds) for the Vox party? Almost three times more likely to vote for the Vox party than to vote for the PP party if you are a man
:::

Which variable have statistically significant coefficients?

Do all variable have significant coefficients for all parties?

What is the effect of the variable gender (in odds) for the Vox party?

------------------------------------------------------------------------

## Select your baseline

If you want a specific category as the baseline, you need to relevel your variable.

```{r}
model_data$party<-relevel(model_data$party,ref = "Vox")
```

------------------------------------------------------------------------

## Analysis with new reference category

```{r}
model2<-multinom(party~agea+eduyrs+gndr,
                data=model_data,
                Hess=TRUE)

stargazer(model2,type="text")

modelsummary(model2,
             stars=TRUE,
             stars_method="simple",
             coef_map=c("agea"="Age",
                        "eduyrs"="Years of education",
                        "gndrMale"="Male"),
             title="Multinomial Regression Results",
             output="html",
             shape= model+term~response,
             gof_map=tibble::tribble(~raw,~clean,~fmt,
                                     "AIC","Akaike Inf. Crit.",2,
                                     "BIC","Bayesian Inf. Crit.",2,
                                     "logLik","Log-Likelihood",2,
                                     "deviance","Deviance",2),
             exponentiate = TRUE
             ) 
```

------------------------------------------------------------------------

## Predictive probabilities

::: r-fit-text
To compute the predicted probabilities, you can use the `ggpredict()` function (as seen in session 2) or you can use the `marginaleffects` package (and more specifically the `plot_predictions()` function). This will directly compute the prediction (you do not need to compute them manually).

`plot_predictions(model, type="probs", condition=variable, draw=FALSE)`:

-   model: name of your multinomial model

-   type="probs": specify that you want to estimate probabilities (so a value between 0 and 1)

-   condition=variable: specify the independent variable for which you want to compute the predicted probabilities

-   draw=FALSE: to get a data frame with the prediction if draw=TRUE, you will also get a ggplot2 graph.
:::

------------------------------------------------------------------------

## Predictive probabilities for age

```{r}
library(marginaleffects)
pred_agea<-plot_predictions(
  model1,
  type="probs",
  condition="agea",
  draw=FALSE)

head(pred_agea)


```

------------------------------------------------------------------------

## Plot predictive probabilities for age

```{r}
pred_agea |>
  ggplot(aes(x=agea,y=estimate,group=group,color=group))+
  geom_line()+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  facet_wrap(~group)+
  theme_light()
```

------------------------------------------------------------------------

## Predictive probability for gender

```{r}
pred_gndr<-plot_predictions(
  model1,
  type="probs",
  condition="gndr",
  draw=FALSE)

pred_gndr |>
  ggplot(aes(x=gndr,y=estimate,group=group,color=group))+
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high))+
  facet_wrap(~group)+
  theme_light()
```

------------------------------------------------------------------------

## Goodness of fit

### Pseudo $R^2$

::: r-fit-text
In multinomial regression, you should compute McFadden's Pseudo $R^2$ which can be interpreted similarly than the $R^2$ (which cannot be computed for a multinomial regression). To compute this pseudo $R^2$, you can use the `pR2()` function from the `pscl` package.

```{r}
library(pscl)
pR2(model2)
```
:::

------------------------------------------------------------------------

### AIC (Akaike Information Criterion)

The AIC criterion is a criterion that should be used when you compare different models. With only one model, you cannot interpret your AIC. The reason behind this is that there are no thresholds for the AIC model to say that is a bad or acceptable model. The AIC is used when you compare nested models. You should select the model with the smallest AIC because a smaller AIC means that the model fits the data better than the model with a higher AIC.

------------------------------------------------------------------------

```{r}
modelA<-multinom(party~agea+eduyrs+gndr,
                data=model_data,
                Hess=TRUE)

modelB<-multinom(party~agea+gndr,
                data=model_data,
                Hess=TRUE)

AIC(modelA, modelB)
```

------------------------------------------------------------------------

### BIC (Bayesian Information Criterion)

Another goodness of fit indicator is the Bayesian Information Criterion. It works similarly to the AIC meaning that you used it to select between models and that you prefer the model with the lowest BIC. The difference between the AIC, is that the BIC will also penalise a model that includes variables that are less useful in your model (to limit problems of overfitting).

```{r}
BIC(modelA,modelB)
```

------------------------------------------------------------------------

# Exercises

------------------------------------------------------------------------

## Exercise 1

::: r-fit-text
We are interested in understanding why people vote Populist Radical-Right parties (this example is inspired by [Luis Sattelmayer](https://luissattelmayer.github.io/quant-2025/session5/session5.html)). We will focus on Germany. We will use the 11th round of the ESS. You can download the data here <https://github.com/scpo-quantimethods/quantitative_methods_two/raw/main/exercices/exercices3/data/ESS11e04_1.csv> and the codebook [here](https://github.com/scpo-quantimethods/quantitative_methods_two/raw/main/exercices/exercices3/data/ESS11e04_1%20codebook.html).
:::

------------------------------------------------------------------------

## Exercise 1 {.scrollable}

1.  Import the data into R. Hint: use `import()` function.

```{r}
ess<-import("https://github.com/scpo-quantimethods/quantitative_methods_two/raw/main/exercices/exercices3/data/ESS11e04_1.csv")
```

2.  Select only the variable `cntry, prtvgde1, imueclt, eduyrs, hinctnta, stflife, trstplt, gndr`and filter the observation to keep only German respondents. Hint: `select` and `filter` function.

    ```{r}
    ess<-ess%>%
      select(cntry, prtvgde1, imueclt, eduyrs, hinctnta, stflife, 
             trstplt, gndr) %>%
      filter(cntry=="DE")
    ```

3.  Have a look at your data. Hint: `glimpse` or `head` function.

    ```{r}
    glimpse(ess)
    head(ess)
    ```

4.  Clean your data.

    -   Transform observations that have are equal to 55, 66, 77, 88, 99, 7777, 8888, 9999 for the prtvgde1, imueclt, eduyrs, hinctnta, stflife, trstplt variable as NAs.

    -   Recode the gndr variable as Man (gnder=1) and Woman (gnder=2).

    -   Create a variable vote based on the prtvgde1 variable (use the name of the party)

        |  |  |
        |------------------------------------|------------------------------------|
        | 1 | Christian Democratic Union/Christian Social Union (CDU/CSU) |
        | 2 | Social Democratic Party (SPD) |
        | 3 | The Left (Die Linke) |
        | 4 | Alliance 90/The Greens (Bündnis 90/Die Grünen) |
        | 5 | Free Democratic Party (FDP) |
        | 6 | Alternative for Germany (AFD) |
        | 7 | Free Voters (Freie Wähler) |
        | 8 | Grassroots Democratic Party of Germany (dieBasis) |
        | 9 | Party for Labour, Rule of Law, Animal Protection, Promotion of Elites and Grassroots Democratic Initiative (Die PARTEI) |

```{r}
ess<-ess%>%
  mutate(across(prtvgde1:trstplt,
                ~case_when(. %in% 
                             c(55, 66, 77, 88, 99, 7777, 
                               8888, 9999)~NA_integer_,
                           TRUE~.)),
         gndr=case_when(gndr==1 ~"Man",
                        gndr==2~"Woman"))

#ess$vote[is.na(ess$prtvgde1)]<-NA
ess$vote[ess$prtvgde1==1]<-"CDU/CSU"
ess$vote[ess$prtvgde1==2]<-"SPD"
ess$vote[ess$prtvgde1==3]<-"Die Linke"
ess$vote[ess$prtvgde1==4]<-"Bündnis 90/Die Grünen"
ess$vote[ess$prtvgde1==5]<-"FDP"
ess$vote[ess$prtvgde1==6]<-"AFD"
ess$vote[ess$prtvgde1==7]<-"Freie Wähler"
ess$vote[ess$prtvgde1==8]<-"dieBasis"
ess$vote[ess$prtvgde1==9]<-"Die PARTEI"
ess$vote<-as.factor(ess$vote)
  
```

5.  Relevel your party variable so that the AFD party is the baseline.

    ```{r}
    ess$vote<-relevel(ess$vote,ref="AFD")
    ```

6.  We want to understand if thinking that immigrants enrich a country's culture decreases the likelihood of voting for the AFD. Run a multinomial regression with the \``imueclt`\` variable as the predictor. Have a look at the coefficients. Hint: use the \`multinom()\` function.

    ```{r}
    reg1<-multinom(vote~imueclt,
                   data=ess,
                   Hess=TRUE)
    stargazer(reg1,type="text")
    ```

7.  Add control variables (gender and education) to your regression. Does the \``imueclt`\` coefficient changes?

    ```{r}
    reg2<-multinom(vote~imueclt+gndr+eduyrs,
                   data=ess,
                   Hess=TRUE)
    stargazer(reg2,type="text")
    ```

8.  Compare the AIC and BIC of models from points 6 and 7. Hint: use the `BIC()` and `AIC()` functions.

    ```{r}
    AIC(reg1,reg2)
    BIC(reg1,reg2)
    ```

9.  Now, we are interested in seeing the influence of trust in politicians (`trstplt` variable) on voting.

    ```{r}
    reg3<-multinom(vote~trstplt,
                   data=ess,
                   Hess=TRUE)
    stargazer(reg3,type="text")
    ```

10. Add controls variable (gender and education) to your regression.

    ```{r}
    reg4<-multinom(vote~trstplt+gndr+eduyrs,
                   data=ess,
                   Hess=TRUE)
    stargazer(reg4,type="text")
    ```

11. Compute the pseudo- \$R\^2\$.

    ```{r}
    pR2(reg4)
    ```

12. Show the predicted probabilities of the trust variable.

    ```{r}
    prediction<-plot_predictions(reg4,
                                type="probs",
                                condition="trstplt",
                                draw=FALSE)

    ggplot(prediction, aes(x=trstplt, y=estimate,group=group,color=group))+
      geom_line()+
      geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
      facet_wrap(~group)+
      theme_light()

    ```

------------------------------------------------------------------------

## Exercise 2

::: r-fit-text
In this exercise, you will use a dataset on student educational choices (you can find it here <https://github.com/scpo-quantimethods/quantitative_methods_two/raw/main/exercices/exercices3/data/student.RData>). The data contains information on 200 students. The student had the possibility of selecting between three different programs to further their education. The choice was between general, vocational and academic. The goal of this example is to understand which predictors can tell us more about why students decided to choose a specific academic program.
:::

------------------------------------------------------------------------

## Exercise 2 {.scrollable}

1.  Import the data

```{r}
student<-import("https://github.com/scpo-quantimethods/quantitative_methods_two/raw/main/exercices/exercices3/data/student.RData")
```

2.  Have a look at your dataset. Recode the *prog* variable so that the categories are not 0, 1, 2 but general, vocational, academic. Hint: use `glimpse`, `mutate()` and `case_when()`.

```{r}
glimpse(student)

student <- student %>%
  mutate(prog = case_when(prog==0 ~ "General",
                          prog==1 ~ "Vocational",
                          prog==2 ~ "Academic"))
```

2.  Now, run a multinomial regression to explain the influence of socio-economic status on the educational program choice.

```{r}
library(nnet)
model1<-multinom(prog~ses,
                data=student,
                Hess=TRUE)

stargazer(model1,type="text")

```

3.  Look at the ODDS ratio

```{r}
exp(coef(model1))

library(modelsummary)
modelsummary(model1,
             stars=TRUE,
             stars_method="simple",
             coef_map=c("seslow"="Low Socioeconomic status",
                        "sesmiddle"="Middle Socioeconomic status"),
             title="Multinomial Regression Results",
             output="html",
             shape= model+term~response,
             gof_map=tibble::tribble(~raw,~clean,~fmt,
                                     "AIC","Akaike Inf. Crit.",2,
                                     "BIC","Bayesian Inf. Crit.",2,
                                     "logLik","Log-Likelihood",2,
                                     "deviance","Deviance",2),
             exponentiate = TRUE
             ) #
```

4.  Add math, read, write, science and socst as control variables to your regression.

    ```{r}
    model2<-multinom(prog~ses+math+read+write+science+socst,
                    data=student,
                    Hess=TRUE)
    stargazer(model2, type="text")
    ```

5.  Look at the predicted probabilities of the socio-economic status

    ```{r}
    pred<-plot_predictions(model2,
                           type="probs",
                           condition="ses",
                           draw=FALSE)

    ggplot(pred, aes(x=ses, y=estimate,group=group,color=group))+
      geom_pointrange(aes(ymin = conf.low, ymax = conf.high))+
      facet_wrap(~group)+
      theme_light()
    ```

6.  What is the goodness of fit of your model?

    ```{r}
    pR2(model2)
    ```
