---
title: "Session 4: difference-in-difference"
format: revealjs
execute:
  echo: true
editor: visual
---

## Last sessions

How to write a research paper

Multinomial regression

------------------------------------------------------------------------

## This week

Rmarkdown

Difference-in-Difference

Homework

------------------------------------------------------------------------

# Rmarkdown

Demo

------------------------------------------------------------------------

# Difference-in-Difference

------------------------------------------------------------------------

## Panel Data

What is panel data?

Is there something specific you need to do when using panel data in regression?

------------------------------------------------------------------------

## Difference-in-Difference

What is it?

When do you use it?

Which type of data do you need?

------------------------------------------------------------------------

## Difference-in-Difference

-   Causal inference

-   Regression technique

-   Compare two groups

-   Pre and Post treatment

------------------------------------------------------------------------

## Difference-in-Difference

$$
Y= \beta_0+\beta_1 Treatment+ \beta_2 Time+ \beta_3 Treatment*Time +\epsilon
$$

-   The effect of the treatment is the interaction effect

------------------------------------------------------------------------

## in R

::: r-fit-text
`lm(formula, data)` :

-   fomula: Y \~ Time\*Treatment (+controls)

-   data: name of the dataset

`feols(formula|FE,cluster=~group,data)`:

-   fomula: Y \~ Time\*Treatment (+controls)

-   FE: fixed effects

-   cluster=\~group: specify that you want clustered SE (if you do not add this argument, you will have IID SE like in lm)

-   data: name of the dataset

With feols function, the stargazer function does not work. You need to use etable().
:::

------------------------------------------------------------------------

## Example

Based on the data from Angelucci and Cagé, 2019.

We are interested in understanding how the introduction of advertisements in TV programs in France has impacted the revenues of newspapers throughout France. The hypothesis being that it impacted only national newspapers and not local ones.

------------------------------------------------------------------------

## Data

You can find the data [here](https://www.openicpsr.org/openicpsr/project/116438/version/V1/view?path=/openicpsr/116438/fcr:versions/V1/data/dta/Angelucci_Cage_AEJMicro_dataset.dta&type=file). You will need to sign in to be able to download it.

```{r}
library(haven)
library(tidyverse)
newspapers <- read_dta("data/Angelucci_Cage_AEJMicro_dataset.dta")

newspapers <-
  newspapers |>
  select(
    year, id_news, after_national, local, national, ra_cst, ps_cst, qtotal
    ) |> 
  mutate(ra_cst_div_qtotal = ra_cst / qtotal, 
         after_national =  if_else(year > 1967, 1, 0),
         after_national_placebo = if_else(year >= 1966, 1, 0),
         across(c(id_news, local, national, after_national), as.factor),
         year = as.factor(year)) 
```

------------------------------------------------------------------------

## Model 1

The basic model

```{r}
did <-
  lm(log(ra_cst) ~ national*after_national,
     data = newspapers)
summary(did)
```

------------------------------------------------------------------------

## Model 2

```{r}
library(fixest)
did2<-feols(log(ra_cst) ~ national*after_national,
            cluster=~id_news,
     data = newspapers)
etable(did2)
```

------------------------------------------------------------------------

## Model 3

```{r}
did3<-feols(log(ra_cst) ~ national*after_national|year+id_news,
            cluster=~id_news,
     data = newspapers)

etable(did3)
```

------------------------------------------------------------------------

## Comparing models

```{r}
did1 <-
  feols(log(ra_cst) ~ national*after_national,
     data = newspapers)
etable(did1,did2,did3)
```

------------------------------------------------------------------------

## Parallel assumptions

The difference-in-difference design is only valid if the parallel trend assumption is valid. The parallel trend assumption is the assumption that before treatment, both the control and treatment had a parallel trend for the outcome variable throughout time. If the parallel trend is wrong, the construction of the counter-factual is wrong, which would create a bias estimator.

------------------------------------------------------------------------

## Graphical check

```{r}
avg_trends <- aggregate(ra_cst ~ national + year, newspapers, mean)
library(ggplot2)
ggplot(avg_trends,aes(x=year,y=ra_cst,color=national))+
  geom_point()+
  ylab("Revenues")+
  theme_bw()
```

------------------------------------------------------------------------

## Placebo check

```{r}
did_placebo <-
  feols(log(ra_cst) ~ national * after_national_placebo | year + id_news,
        cluster=~id_news,
     data = newspapers)
etable(did_placebo)
```

------------------------------------------------------------------------

# Exercises

------------------------------------------------------------------------

## Exercise 1

In the USA, in 2000 a law was passed with the Right To Carry (RTC) arms in certain states (simulated example). US lawmakers are interested in knowing if this increased or not crimes in cities.

------------------------------------------------------------------------

## Exercise 2

In 2021, a few local governments implemented policies with stricter emissions controls, other local governments have not (simulated example). The air quality data is available from 2018 to 2022.

------------------------------------------------------------------------

## Exercise 3

In 2019, some cities decided to introduce a smoking ban on the terraces of bars and restaurants (simulated example). We want to analyse the effect of the smoking ban on the number of hospitalisations linked to respiratory disease. Use the smoking dataset.
