---
title: "Exercises 6 - Time Series"
format: html
editor: visual
execute:
  echo: false
  eval: false
---


# Exercise 1

In this exercise, we are interested in the relationship between GDP and unemployment (simulated dataset). You can find the dataset here <https://github.com/scpo-quantimethods/quantitative_methods_two/raw/main/exercices/exercices6/data/panel.Rdata>

1. Download the data.

```{r}
library(rio)
#panel<-import("https://github.com/scpo-quantimethods/quantitative_methods_two/raw/main/exercices/exercices6/data/panel.Rdata")
load("panel.Rdata")
```


2. Create a new dataset where you only keep the observations from Japan.

```{r}
library(tidyverse)
japan<-subset(panel,country=="Japan" )
```

3. Plot the time series of unemployment and GDP

```{r}
unemp<-ts(japan$unemp,start=first(1990), frequency=1)
plot.ts(unemp)

gdp<-ts(japan$gdp,start=first(1990), frequency=1)
plot.ts(gdp)
```

4. Are the two times series stationary?

```{r}
library(tseries)
adf.test(gdp)


adf.test(unemp)
unemp_diff<-diff(unemp)
adf.test(unemp_diff)
```


4. Identify, estimate and diagnose the GDP time series.

```{r}

acf(gdp)
pacf(gdp)
forecast::auto.arima(gdp)

mod1<-arima(gdp,order=c(1,0,0))
mod1

```


5. Identify, estimate and diagnose the democratization time series.

```{r}

acf(unemp)
pacf(unemp)
forecast::auto.arima(unemp)

mod2<-arima(unemp,order=c(0,0,1))
mod2

```

6. Does unemployment decrease GDP ? Check with the prewhitening technique.

```{r}
unemp.res<-mod2$residuals
adf.test(unemp.res)
gdp.res<-mod1$residuals
adf.test(gdp.res)
ccf(gdp.res,unemp.res,lag.max=10)

```


7. Does unemployment decreases GDP? Check with the Granger causality technique.

```{r}
library(lmtest)
grangertest(gdp~unemp,order=5,data=japan)
grangertest(unemp~gdp,order=5,data=japan)
```


8. Now, we are interested in doing a TSCS analysis, so we will use the whole data set again. Run a TSCS analysis with fixed effects. Use the fixed effects estimator

```{r}
library(plm)
fe<-plm(
    gdp ~ unemp ,
    data = panel,
    index = c("country", "year"),
    model = "within"
  )
```


9. Run a TSCS analysis with random effects.

```{r}
re<-plm(
    gdp ~ unemp ,
    data = panel,
    index = c("country", "year"),
    model = "random"
  )
```


10. Run a Hausman test. Is it better to use fixed or random effects in this case?

```{r}
phtest(fe, re)
```


11. Run a PCSE with lagged dependent variable model.

```{r}
library(tidyverse)

library(pcse)
panel <- panel %>%
  group_by(country) %>%
  mutate(l.gdp = dplyr::lag(gdp))

panel2<-na.omit(panel)

panel2$country <- as.factor(panel2$country)
panel2$year <- as.numeric(as.character(panel2$year))

ols<-lm(gdp~l.gdp+unemp,data=panel2)
pcse<-pcse(ols, groupN=panel2$country, groupT=panel2$year)
summary(pcse)
```



# Exercise 2

In this exercise, we are using the data from Abou-Chadi ([2016](https://www.cambridge.org/core/journals/british-journal-of-political-science/article/niche-party-success-and-mainstream-party-policy-shifts-how-green-and-radical-right-parties-differ-in-their-impact/AEF265EED2CAA81C35A3808F47F5908A#article)) article. You can download the data on the supplementary materials page of the paper. The example is inspired by [Luis Sattelmayer](https://luissattelmayer.github.io/quant-2025/session6/session6.html#hierarchical-models). We are interested in understanding the effect of increasing radical right votes on mainstream parties' shift to anti-immigrant positions.

1. Download the data and import it into R. Select the variable countryname, cultposition, sumright, rileright, limmig, govparty pervote, partyname, date

```{r}
library(tidyverse)
dt <- haven::read_dta("abouchadi_data.dta") 


data<-dt %>%  select(
    countryname,
    cultposition,
    sumright,
    rileright,
    limmig,
    govparty,
    pervote,
    partyname,
    date
  )

```


2. Run a TSCS analysis with fixed effects. Use the fixed effects estimator

```{r}
library(plm)
fixed_effects <-
  plm(
    cultposition ~ sumright ,
    data = data,
    index = c("partyname", "date"),
    model = "within"
  )
summary(fixed_effects)
```


3. Run a TSCS analysis with fixed effects. Use the dummy variable estimator

```{r}
fixed_effects2<-lm(cultposition~sumright+factor(partyname)-1,data)
summary(fixed_effects2)
```


4. Run a TSCS analysis with random effects.

```{r}
random_effects <-
  plm(
    cultposition ~ sumright ,
    data = data,
    index = c("partyname", "date"),
    model = "random"
  )

summary(random_effects)
```

5. Run a Hausman test. Is it better to use fixed or random effects in this case?

```{r}
phtest(fixed_effects, random_effects)
```

6. Redo points 3 and 4 but a lagged dependent variable into the regressions.

```{r}
data <- data %>%
  group_by("partyname") %>%
  mutate(lag_cultposition= dplyr::lag(cultposition))

data2<-na.omit(data)

fixed_effects3<-lm(cultposition~sumright+lag_cultposition+factor(partyname)-1,data2)
summary(fixed_effects3)

random_effects2 <-
  plm(
    cultposition ~ sumright + lag_cultposition ,
    data = data2,
    index = c("partyname", "date"),
    model = "random"
  )

summary(random_effects2)
```

