---
title: "Exercices 2 : logistic regressions"
format: html
---

## 1. Turnout and rain

You are interested in knowing whether turnout, the participation in elections, is conditional on weather: that is, do people vote less when it rains? To do so, you are provided with a (fictitious) data set on American elections in 2012, 2016, and 2020.

0. Load the following packages : 
```{r}
library(ggeffects)
library(ggplot2)
library(tidyverse)
library(rio)
```
1. Using `rio` and the function `import`, import the data from <>. 
2. Have a look at your data using `glimpse()`. Drop all variables except `voted, voted_prev, rain_mm, age, female, year`.
3. You want to inspect whether the chances of voting change with the rainfall on the day of the election, `rain_mm`. The variable `vote` tells if people have voted in the elections. What would be your dependent and independent variables ? Are they categorical or continuous ?
4. Count the total number of voters and summarize the `rain_mm` variable. Are there NA's in the data set ?
5. Using `mutate`, `case_when`, prepare the `voted` variable for the logistic regression. 
6. Assuming you are estimating a logistic regression, what is the mathematical expression of the probability function as a function of rainfall, denoted $R_mm$ ? What is the corresponding expression for the logit of that probability? Finally, what is the expression for the odds ratio?
7. Run a logistic regression with one variable to estimate whether the rain changes anything in the probabilities of voting. For a 1-mm increase in the rainfall, how does the likelihood of voting change ? How significant is that result ? 
8. Does voting in the previous election change the probabilities of voting in the next one ?
9. Add the variables `voted_prev, female, age, rain_mm, as.factor(year)` in the predictors, and compare the three models, using the `stargazer` package;
10. Use the `ggpredict` function from package `ggeffects` to predict the probabilities of voting depending on rain_mm. Have a look at your predictions.
11. Use `ggplot` to plot the predicted probabilities with their confidence intervals, with their confidence intervals. Hint : use `+ geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2)` to display smooth confidence intervals, and `geom_line()` to add a line connecting the dots.

```{r}
#| echo: false
#| eval: false

# run install.packages() if you haven't the packages already installed
library(ggeffects)
library(ggplot2)
library(tidyverse)
library(rio)
library(stargazer)
#1.1
rain_vote <- import("/Users/paulservais/Library/CloudStorage/ProtonDrive-paservai@protonmail.com-folder/Website/quantitative_methods_two/exercices/exercices2/data/rainfall.dta")

#1.2
glimpse(rain_vote)
rain_vote <- rain_vote %>% select(voted, voted_prev, rain_mm, age, female, year)

#1.3
# DV : voting (binary)
# IV : raining. Likely exogeneous (uncorrelated with the error)

#1.4
table(rain_vote$voted)
summary(rain_vote$rain_mm)
summary(rain_vote)

#1.5
# with case_when()
rain_vote <- rain_vote %>% mutate(
  voted = case_when(
    voted == "voted" ~ 1,
    voted == "did not vote" ~ 0
  )
)

#1.6
log_model <- glm(
  data = rain_vote,
  voted ~ rain_mm,
  family = "binomial"
)
exp(log_model$coefficients)-1
exp(coef(log_model))-1

#1.7
log_model2 <- glm(
  data = rain_vote,
  voted ~ voted_prev,
  family = "binomial"
)
exp(log_model2$coefficients)-1
exp(coef(log_model2))-1

#1.8
log_model3 <- glm(
  data = rain_vote,
  voted ~ voted_prev+female+age+rain_mm+as.factor(year),
  family = "binomial"
)


#1.8
stargazer(log_model, log_model2, log_model3,
  type = "text",
  dep.var.labels = "Voting Behavior",
  dep.var.caption = c("Voting turnout; 0 = abstention | 1 = voted"),
  star.cutoffs = c(0.05, 0.01, 0.001)
)

#1.9
predicted_probs <- ggpredict(log_model3, terms = "rain_mm") 
predicted_probs 

#1.10
ggplot(predicted_probs, aes(x = x, y = predicted)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +  # smooth CI
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(shape = 21, size = 3, color = "steelblue") +
  labs(
    title = "Predicted probability of voting by rainfall",
    x = "Rainfall (mm)",
    y = "Predicted probability of voting"
  ) +
  theme_minimal()
```

## 2. ESS : voting in Italy
You are provided with a simplified dataset containing ESS data for Italy.

1. Download the dataset and import it into your local environment.
2. Inspect the dataset using `glimpse`.
3. Examine the values of the vote variable and modify them using `case_when` if necessary. Prepare the variable for a logistic regression. Review the other variables as well and modify them as needed. You may draw on the theory section for this session for guidance.
4. Run a logistic regression to analyze whether the number of years of education influences the probability of voting.
5. Create a variable `age` and include the variables `polintr, trstplt, clsprty, gndr, age, eduyrs` in the logistic regression.
6. Compare the two models using `stargazer`.
7. Plot the predicted probability of voting by years of education.

```{r}
#| echo: false
#| eval: false


#1.1
ess <- import("/Users/paulservais/Library/CloudStorage/ProtonDrive-paservai@protonmail.com-folder/Website/quantitative_methods_two/sessions/session2/data/ess_it.csv")

glimpse(ess)

#1.2
table(ess$vote)
ess <- ess |> 
  # mutating the dependent variable to get rid of any unnecessary rows
  # as we have done above
  mutate(
    vote = case_when(
      vote %in% c(3, 7, 8, 9) ~ NA_integer_,
      vote == 2 ~ 0,
      vote == 1 ~ 1, 
      TRUE ~ vote
    ),
    # Recode 'polintr' to drop values 7 to 9
    polintr = case_when(
      polintr %in% c(7:9) ~ NA_integer_,
      TRUE ~ polintr
    ),
    # Recode 'clsprty' 1/2 into 0/1 and drop 7, 8
    clsprty = case_when(
      clsprty == 1 ~ 0,
      clsprty == 2 ~ 1,
      clsprty %in% c(7, 8) ~ NA_integer_,
      TRUE ~ clsprty
    ),
    # Recode 'gndr' into 0/1 and handle other cases as NAs
    gndr = case_when(
      gndr == 1 ~ 0,
      gndr == 2 ~ 1,
      TRUE ~ NA_integer_
    ),
    # recode eduyrs
    eduyrs = case_when(
      eduyrs %in% c(1:55) ~ eduyrs,
      TRUE ~ NA_integer_
    )
  )
#1.3
table(ess$vote)

#1.3
logit1 <- glm(
  vote ~  eduyrs,
  data = ess %>% mutate(age = 2026-yrbrn),
  family = "binomial")

logit2 <- glm(
  vote ~ polintr + trstplt + clsprty + gndr +age + eduyrs,
  data = ess %>% mutate(age = 2026-yrbrn),
  family = "binomial")

#1.4
# simply use exp() on the coefficients of the logit
exp(coef(logit))
# here would be a second way of doing it
exp(logit$coefficients)
exp(cbind(
  OR = coef(logit), 
  confint(logit)))

#1.5
stargazer(logit1, logit2, 
          type = "text",
         dep.var.labels = "Voting Behavior",
         dep.var.caption = c("Voting turnout; 0 = abstention | 1 = voted"),
          star.cutoffs = c(0.05, 0.01, 0.001))

#1.6
predicted_probs <- ggpredict(logit, terms = "eduyrs")
predicted_probs

ggplot(predicted_probs, aes(x = x, y = predicted)) +
  geom_line(color = "blue", size = 1) +
  geom_point() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "grey", alpha = 0.3) +
  labs(
    title = "Predicted Probability of Voting by education years",
    x = "Education years",
    y = "Predicted Probability of Voting"
  ) + 
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

```

